---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

# hectorpractice

<!-- badges: start -->
<!-- badges: end -->

The `hectorpractice` package can be used to run Hector in an interative process with the capability of computing probabilistic outcomes of climatic variables in provided year ranges for different climate impact scenarios. 

## Installation

You can install the development version of `hectorpractice` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ecolo-joe/hectorpractice")
```

## Example

This is a basic example which shows you how to use the iterative_hector function to run Hector 10 times for a provided emission scenario to produce a data frame of median global average temperature increase (variable = global_tas) for the years 2000-2100:

```{r example}
library(hectorpractice)
```

First, an emission scenario must be loaded and stored in a core to initialize Hector. Scenarios are stored in the `hector` package in INI config format.

```{r cars}
# Load ini file 
ini_file <- system.file("input/hector_ssp245.ini", package = "hector")

# Build core with ini file
ssp245 <- hector::newcore(ini_file)
```

Once a core is constructed it can be passed to the `iterative_hector` function. The user will also need to supply other required arguments.

```{r}
# Identify climatic variables and years of interest
variable <- c("global_tas")
year_range <- 2000:2100

# Use core to run iterative_hector
results <- iterative_hector(ssp245,
                            op = median,
                            var = variable,
                            years = year_range,
                            runs = 10)

# View results 
head(results)

```

The result of `iterative_hector` is a data frame filtered to only include the year range specified by the user for the variables specified by the user. The data frame also includes the parameter values used for each hector run (based on random draws from normal distribution). Parameter values are re-drawn from the normal distribution for each run. Finally, the `metric` column provides the median `global_tas` for each run and the `run_number` informs the user of which hector run they are observing values and metrics from.   
