---
title: "hectorpractice-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hectorpractice-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Give a good description of hector practice -- what is it doing? include the capability of running hector through an iterative process using randomly generated parameters for each iteration which provides a level of uncertainty to the model we did not previously have. In addition, the package gives the capability to calculate metrics that can summarize climate variable projections for year ranges. For example, it can compute projections of mean global surface temperature warming by the end of the century. Last, it will have the ability to provide probabilities of such climatic variable projections. But make this all sound better.

This vignette will run through the basic functionality of this package. First, it shows how to produce a data frame of parameter values drawn from randomly from normal distributions. Next, it will use the parameters produced to run Hector in an iterative process, whereby parameter values are changed with each subsequent model run. Then it shows how to use an iterative Hector output to compute mean global surface temperature warming for each iteration. 

eventually there will be probability calculation information in here somewhere.

####

# Set-up

## Configure Hector core

First, load the `hectorpractice` package 

```{r setup}
library(hectorpractice)
```

Next, we will initialize a core or Hector instance. While some functions in this package do not require a core input, we will need one to run the Hector model. 

More information about establishing a new core for running Hector can be found in the tutorial for using the [Hector R interface](https://jgcri.github.io/hector/articles/intro-to-hector.html).

```{r}
# Reading INI file with emission scenario of interest
ini <- system.file("input/hector_ssp245.ini", package = "hector")

# Creating a new core 
c_ssp245 <- newcore(ini)

```

This should configure a new Hector core and add it to the Global Environment.
--> Should I give more information about the core? Can add but will likely be replicate explanations that can also be found the the tour of Hector vignette.

A core is an object that contains information about Hector inputs and outputs.

Now that we have a core, we can generate parameter values that can be subsequently set in the core for a series of Hector runs.

## Generate parameter values

We can generate a data frame of parameter values using `generate_params()`. This function will randomly sample a normal distribution of parameters and store those values in a data frame. We can specify how many random draws we would like to complete, which will also be setting the number of times we will run Hector.

```{r}
# Generate a data frame of random parameter values
param_values <- generate_params(10)
head(param_values)
```

We now have a data frame object called `param_values` that contains our model parameters with 10 unique values that were sample for each parameter.

We can use this parameter data frame to run Hector through an iterative process, setting new parameter values with each run until Hector outputs have been produced for each set of parameter values we generated.  

####

# Running Hector with iterations

Now that we have configured a Hector core and generated a series of parameter values, we can run Hector through iterative model runs using the `iterative_hector()` function.

`iterative_hector()` requires the some specific information in order to produce the output we want. As previously mentioned, it requires a core and a data frame of parameter values. In addition it requires a specification of the variables and year ranges to fetch from the core object. Below we will specify that we want to retrieve results for global average surface temperature (global_tas) for the years 2000 to 2300. We create an object called `results` which will store our output in a data frame.

```{r, message=FALSE}
# Run Hector model repeatedly over all parameter values
results <- iterative_hector(c_ssp245, var = "global_tas", years = 2000:2300, params = param_values)
head(results)
```

Viewing the full results will show that global average surface temperature values from 2000-2300 for 10 separate runs, which is indicated by `run_number`. Each run uses different parameter estimates to complete the model. 

The result is a `data.frame` and therefore we can plot the results easily in `ggplot2`.

```{r}
# plotting results from iterative Hector run
library(ggplot2)

ggplot(data = results, aes(x = year, y = value, group = run_number)) +
  geom_line()
```

From this result we can see how incorporating uncertainty in parameters included in the model influences trajectories of climate variables, in this case global average surface temperature. This shows the non-deterministic result of climate model projections when parameter uncertainty is included in Hector model runs. 

####

**What would be next? -- computing metrics and using the output to conduct probability of variable outcomes?**

**Running multiple scenarios? plotting max, min, median of the parameter uncertainty?**
