---
title: "matilda-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{matilda-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

matilda (NAME!?) is a package that provides probabilistic capability to the Hector simple climate model. This package gives the user the ability to randomly generate model parameters from a normal distribution, run Hector using the generated parameters, screen model outputs for realism, and compute probabilities of climate variables projections. This package allows Hector to automate the incorporation of parameter uncertainty. Including parameter uncertainty allows Hector to run multiple times across a range of possible parameter values. into and quantifies the likelihood of future climate change outcomes.

In addition, the package has functions available that allow the user to identify information they are interested in fetching and computing from Hector runs. Here we define a metric as a combination of identifiers that help fetch data and calculate a "metric value" from a Hector output. For example, it is possible to compute the probability that mean global surface warming will remain between 2-3$^\circ$ C by the end of the century (2100). In this case the metric could be identified as:

`Probabilistic Hector Metric:  mean global_tas 2000  to  2100`

Once the model is run, the user can screen outputs using observed data to identify the most realistic model results (those that most closely align with observed data). We call this a scoring "criterion". `matilda` has pre-established criterion for scoring Hector runs based on proximity to observed data, but also is capable of allowing the user to establish their own scoring criterion.  

This vignette will work through the basic functionality of the matilda (NAME!?) package. First, we will show how to produce a data frame of parameter values drawn randomly from normal distributions of important model parameters. Next, we will use the parameter values produced to run Hector in an iterative process, whereby parameter values are substituted (without replacement) for multiple model runs. Then, we shows how to use an iterative Hector output to compute a climatic variable metric for each Hector run. These metrics will then be organized into bins and probabilities will be calculated for the projection of each climatic variable.

As a final simple analysis example, we will show how these functions give Hector the capability to be used as a coupled climate model to compute probabilistic projections of future climate change using emissions derived from two different Shared Socio-economic Pathways (SSPs). 

####

# Set-up

## Configure Hector core

First, load the `matilda` package 

```{r setup}
library(matilda)
```

Next, we will initialize a "core" or Hector instance, we will need this core to run the Hector model. 

More information about establishing a new core for running Hector can be found in the tutorial for using the [Hector R interface](https://jgcri.github.io/hector/articles/intro-to-hector.html).

```{r}
# Reading INI file with emission scenario of interest
ini <- system.file("input/hector_ssp245.ini", package = "hector")

# Creating a new core 
c_ssp245 <- newcore(ini)

```

This new Hector core is an object that contains information about Hector inputs and outputs.

Now we can generate parameter values. The parameter values that we produce will be subsequently used to set parameter variables in our core for a series of Hector runs.

## Generate parameter values

We can generate parameter values using `generate_params()`. This function randomly samples a normal distribution of each parameter and stores sampled parameter values in a data frame. We need to specify how many random draws of the parameter distributions we would like to complete. The number of sample draws we indicate will also be the number of times we will run Hector.

```{r}
# Generate a data frame of random parameter values
param_values <- generate_params(c_ssp245, 10)

head(param_values)
```

We now have a data frame object called `param_values`. This data frame contains our model parameter names with 10 unique values that were sampled from a normal distribution of each parameter.

We can use this parameter data frame to run Hector through an iterative process. `matilda` can run Hector with iterations that will run Hector multiple times, setting new parameter values with each run until an output has been produced for each set of parameter values generated.  

####

# Running Hector with iterations

Once we have configured a Hector core and generated a series of parameter values, we can complete iterate runs in Hector using the `iterate_hector()` function.

`iterate_hector()` requires some specific information in order to produce an output we want. As previously mentioned, it requires a core (arg = `core`) and a data frame of parameter values (arg = `params`). In addition, it requires the user to indicate which variables and year ranges to fetch from the core object (arg = `metric`). 

`matilda` has the ability to store desired metric identifiers in an object that can be easily passed to many different functions in the package, including `iterate_hector`. 

We will use the `new_metric` function to define our metric and use that object as our `metric` argument in `iterate_hector`.

Below we will create a metric indicating that we want to retrieve results for global average surface temperature (global_tas) for the years 2000 to 2300. We will also specify that we are interested in computing mean global_tas data. It is important to note that the operation stored in our metric will **not** be used in the `iterate_hector` function, but will be important for subsequent steps in the analysis. 

For this example we will be using observed atmospheric CO2 as a scoring criterion. This criterion is easy to use because it is already defined within the `matilda` package. In order to screen our Hector runs based on observed CO2, we will need to the atmos_co2 variable included in our Hector output. Therefore, we will include it as a desired variable when defining our metric.  

```{r}
# Setting a new_metric object for Hector analysis
my_metric <- new_metric(c(GLOBAL_TAS(), ATMOSPHERIC_CO2()), 2000:2300, median)

print(my_metric)
```

Now that have all of the required arguments for `iterate_hector` we can run the model. We will create an object called `results` which will store our output as a data frame.

```{r, message=FALSE}
# Run Hector model repeatedly over all parameter values
results <- iterate_hector(core = c_ssp245, metric = my_metric, params = param_values)

head(results)
```

Viewing the full results will show that global average surface temperature values from 2000-2300 for 10 separate runs, which is indicated by `run_number`. As a reminder, each run uses different parameter estimates to complete the model. 

The result is a `data.frame` and therefore we can plot the results easily in `ggplot2`.

```{r}
# plotting results from iterative Hector run
library(ggplot2)

ggplot(data = results[results$variable %in% "global_tas", ], 
       aes(x = year, y = value, group = run_number)) +
  geom_line() +
  facet_wrap(~variable, scales = "free_y")
```

From this result we can see how incorporating parameter uncertainty in Hector influences trajectories of climate variable projections, in this case global average surface temperature. We can view this as a non-deterministic result of climate model projections when parameter uncertainty is included in the Hector model.

Once we have our Hector result, we can begin to address broader questions about climate projections. For example, we may be interested in what the median global_tas values are from 2000-2100 for each model run, and what the probability is that median warming of the global average surface temperature will remain below 2$^\circ$ C by 2100?

####

# Calculating metrics and probabilistic outcomes

After producing a result from `iterate_hector` we can calculate our metrics of interest for each Hector run. As a reminder, we have already specified a metric we are interested in (`my_metric`). This metric object includes outputs from 2000-2300. 

But if we are now interested in a reduced year range (2000-2100) we can just as easily create a new metric specific to those goals.

```{r}
# New metric object to address question of warming below 2 C by 2100
#my_metric2 <- new_metric("global_tas", 2000:2100, median)

#print(my_metric2)
```

With the new metric object and our `iterate_hector` output data frame, we can utilize the `metric_calc` function to calculate our metric (median global_tas from 2000 to 2100) for each of our Hector runs.

```{r}
# Getting results for our metric of interest
#metric_result <- metric_calc(results, my_metric2)

#head(metric_result)
```

We can use the `metric_calc` results to organize data into bins for which we define the ranges. 

```{r}
#binned_metrics <- bin_metrics(metric_result, c(1, 1.5, 2.0, 2.5, 3.0, 4.0))

#print(binned_metrics)

```

#### EDIT 

After data are binned we can use the table result to calculate the probability of specific outcomes of the climate varaible we are interested in. 

In this example... 

```{r}
#prob_result <- prob_calc(binned_metrics)

#print(prob_result)
```

